Black-hole ringdown as a probe of higher-curvature gravity theories

### 2022-03-08 papers/paper_alt_theor_bounds/paper_alt_theor_bounds.tex
hs: making notation a bit uniform across sections

The GW signal from a BBH merger can be determined in GR by a unique set of parameters -- the masses and spins of the two BHs, $(m_1,\, m_2,\, \vec{s}_1,\, \vec{s}_2)$, the sky location determined by the luminosity distance $d_L$, right ascension $\alpha$ and declination $\delta$, and the orientation of the binary given by the inclination and polarisation angles, $(\iota, \psi)$. The set is completed by the choice of a reference time and phase, $(t_0,\, \phi_0)$. If we further assume that the spins of the individual BHs are restricted to be parallel to the orbital angular momentum, then we reduce the 6 components of spin to just 2, $(\chi_1, \chi_2)$ and our entire parameter set from 15 to 11. We define some additional parameters and set some conventions that will be useful in our analysis later, namely, the totat mass $M=m_1+m_2$, the chirp mass $\mathcal {M}=(m_{1}m_{2})^{3/5}/(m_{1}+m_{2})^{1/5}$, the asymmetric mass ratio $q=m_1/m_2$ with the convention $m_1 \geqslant m_2; q \geqslant 1$ and the symmetric mass ratio of the binary, $\nu = m_1m_2/(m_1+m_2)^2$.

The polarisations of the GW signal (in the observer's frame) are,
%
\begin{equation}
h_+(\iota,\varphi_0;t ) - i h_\times(\iota,\varphi_0;t) = \sum_{l, m} {}_{-\!2}Y_{l m}(\iota,\varphi_0)\, h_{l m}(t)\,,
\end{equation}
%
where ${}_{-\!2}Y_{l m}(\iota,\varphi_0)$ are the $-2$ spin-weighted spherical harmonics described by standard angular dependence indices $(l,m)$.\footnote{Note that we use $l$ to denote the angular dependence index and $\ell$ to denote the coupling constant in Sec.~\ref{sec:review_parspec}.} As our baseline model, we use the computationally efficient (time-domain) multipolar waveform model for quasicircular spin-aligned BBH mergers described in \cite{Mihaylov:2021bpf} (henceforth referred to as \SEOB{}\footnote{This waveform model is available in \texttt{LALSuite} \cite{lalsuite} as the \texttt{SEOBNRv4HM\_PA} waveform approximant.}) which contains the modes, $(l, |m|)=(2,2),(2,1)$, $(3,3)$, $(4,4)$, and $(5,5)$~\cite{Cotesta:2018fcv,Mihaylov:2021bpf}. The model uses an effective-one-body approach combined with a post-adiabatic solution of the equations of motion to describe the inspiral-plunge waveform, $h_{l m}^\mathrm{insp-plunge}$. An accurate description of the merger is incorporated through calibration with NR simulations, as described in~\cite{Cotesta:2018fcv}, along with information for the merger and ringdown phases, from BH perturbation theory. The merger-ringdown waveform, $h_{l m}^\mathrm{merger-RD}$, is then stitched to inspiral-plunge waveform, $h_{l m}^\mathrm{insp-plunge}$ at a certain time $t = t^{\textrm{match}}_{l m}$, as
%
\begin{equation}
h_{l m}(t) = h_{l m}^\mathrm{insp-plunge}\, \theta(t_\mathrm{match}^{l m} - t) + h_{l m}^\mathrm{merger-RD}\,\theta(t-t_\mathrm{match}^{l m})\,,
\end{equation}
%
where $\theta(t)$ is the Heaviside step function. The merger-ringdown waveform is expressed as an exponentially damped sinusoid ~\citep{Bohe:2016gbl,Cotesta:2018fcv,Mihaylov:2021bpf}
%
\begin{equation}
\label{RD}
h_{l m}^{\textrm{merger-RD}}(t) = \nu \ \tilde{A}_{l m}(t)\ e^{i \tilde{\phi}_{l m}(t)} \ e^{-i \sigma_{l m 0}(t-t^{\textrm{match}}_{l m})},
\end{equation}
%
where
%
\begin{align}
\sigma_{l m 0} &= {\rm Re}(\sigma_{l m 0}) + i \, {\rm Im}(\sigma_{l m 0})\,,
\nonumber \\
&= \omega_{l m 0} - i / \tau_{l m 0}\,,
\end{align}
%
are the complex frequencies of the fundamental (n=0 overtone) QNMs of the remnant BH. The frequencies $\omega_{l m 0}$ and damping times $\tau_{l m 0}$ of the $(l, m)$ QNM can be read off from the real and imaginary parts of  $\sigma_{l m0}$ respetively. The functions $\tilde{A}_{l m}(t)$ and $\tilde{\phi}_{l m}(t)$ are defined in~\cite{Bohe:2016gbl,Cotesta:2018fcv}.

@@

\subsubsection{Bayesian formalism}
If we assume that a GW signal observed in detector data $d$ is accurately
described by our waveform model \pSEOB{}, we can infer the parameters of the
model, $\bm{\lambda}$, given the hypothesis $\mathcal{H}$, using the Bayes
theorem,
%
\begin{equation}
P(\bm{\lambda} \vert d, \mathcal{H}) =
\frac{P(\bm{\lambda} \vert \mathcal{H}) P(d \vert \bm{\lambda},\mathcal{H})}{P(d \vert \mathcal{H})}\,,
\label{eq:bayes}
\end{equation}
%
where $P(\bm{\lambda} \vert d, \mathcal{H})$ is the posterior probability distribution,
$P(\bm{\lambda} \vert \mathcal{H})$ the prior,
$P(d \vert \bm{\lambda},\mathcal{H})$ the likelihood, and
$P(d \vert \mathcal{H})$ the evidence. The set of parameters, $\bm{\lambda}$, detailed in Sec.~\ref{sec:review_pSEOB}
include the set of beyond-GR parameters, $\{\ell, \delta \omega_k^{(j)}, \delta \tau_k^{(j)}\}$.
%
As discussed in Secs.~\ref{sec:review_theories} and \ref{sec:theory_specific_qnm}, there is a single coupling
constants in the cases of EdGB and dCS theories, but two in the cubic and
quartic EFTofGR. We also hold the other beyond-GR parameters, $\{\delta \omega_k^{(j)}\},$ and $\{\delta \tau_k^{(j)}\}$
fixed to theory-specific predictions. Hence, our set of beyond--GR parameters only includes the parameter $\{\ell \}$.
%

Assuming Gaussian and stationary noise, we can write the (log) likelihood function as,
%
\begin{equation}
\ln P(d \vert \bm{\lambda},\mathcal{H}) \propto
- \tfrac{1}{2}
\langle
d - h(\bm{\lambda}) \vert d - h(\bm{\lambda})
\rangle\,,
\end{equation}
%
with noise-weighted inner product $\langle \cdot | \cdot \rangle$ defined as,
%
\begin{equation}
\langle A | B \rangle =
\int_{f_{\rm low}}^{f_{\rm high}} \, \dd f \,
\frac{\tilde{A}^{\ast}(f) \tilde{B}(f) + \tilde{A}(f) \tilde{B}^{\ast}(f)}{S_{n}(f)}
\end{equation}
%
where $\tilde{A}(f)$ is the Fourier transform of $A(t)$, the asterisk denotes
complex conjugation and $S_{n}(f)$ is the power spectrum density of the
detector.
%
Assuming a specific prior distribution for our parameters, we stochastically
sample over the parameter space using a \texttt{LALInference} Markov-Chain
Monte Carlo algorithm.~\cite{lalsuite}. We subsequently marginalize over the
remaining parameters to obtain the posterior probability distribution
function~(PDF) on $\ell$. An example of this is shown in Fig. XX.

@@

If we find that our posterior obey both conditions~\eqref{eq:eft_bound}
and~\eqref{eq:parspec_bound}, we combine individual posteriors on $\ell$ from
each event.
%
% When possible, we also combine the marginalized posterior distribution on
% $\ell$ for both event GW150914 and GW200129.
%
To do so, we employ one of the approaches of Ref.~\cite{Perkins:2021mhb} which
is relies on first obtaining a kernel density estimator (KDE) as an approximation for
the posterior distribution on $\ell$ for each individual events. Next, we multiply
these posteriors together to obtain the cumulative posterior.
%
(See also Refs.~\cite{DelPozzo:2011pg,Cardenas-Avendano:2019zxd,Carullo:2021dui}.)
%
This method can have difficulties in dealing with hard cut-offs in the
distribution, such as the one we impose at $\ell = 0$. We address this issue as
done in Ref.~\cite{Perkins:2021mhb}, by first doubling our number of samples as
$\{ s_i \} \cup \{ - s_i \}$, i.e., mirroring the samples across $\ell = 0$.
%
We then renormalize the associated KDE numerically in the range $\ell = [0,
\infty]$.

>>>>

The GW signal from a BBH merger can be determined in GR by a unique set of parameters $\bm{\theta}$, that
includes the masses and spins of the two BHs, $(m_1,\, m_2,\, \vec{s}_1,\, \vec{s}_2)$, the sky location determined by the luminosity distance $d_L$, right ascension $\alpha$ and declination $\delta$, and the orientation of the binary given by the inclination and polarization angles, $(\iota, \psi)$. The set is completed by the choice of a reference time and phase, $(t_0,\, \phi_0)$. If we further assume that the spins of the individual BHs are restricted to be parallel to the orbital angular momentum, then we reduce the 6 components of spin to just 2, $(\chi_1, \chi_2)$ and our entire parameter set from 15 to 11. We define some additional parameters and set some conventions that will be useful in our analysis later, namely, the totat mass $M=m_1+m_2$, the chirp mass $\mathcal {M}=(m_{1}m_{2})^{3/5}/(m_{1}+m_{2})^{1/5}$, the asymmetric mass ratio $q=m_1/m_2$ with the convention $m_1 \geqslant m_2; q \geqslant 1$ and the symmetric mass ratio of the binary, $\nu = m_1m_2/(m_1+m_2)^2$.

The polarization of the GW signal (in the observer's frame) are,
%
\begin{equation}
h_+(\iota,\varphi_0;t ) - i h_\times(\iota,\varphi_0;t) = \sum_{l, m} {}_{-\!2}Y_{l m}(\iota,\varphi_0)\, h_{l m}(t)\,,
\end{equation}
%
where ${}_{-\!2}Y_{l m}(\iota,\varphi_0)$ are the $-2$ spin-weighted spherical harmonics described by standard angular dependence indices $(l,m)$.\footnote{Note that we use $l$ to denote the angular dependence index and $\ell$ to denote the coupling constant in Sec.~\ref{sec:review_parspec}.} As our baseline model, we use the computationally efficient (time-domain) multipolar waveform model for quasicircular spin-aligned BBH mergers described in \cite{Mihaylov:2021bpf} (henceforth referred to as \SEOB{}\footnote{This waveform model is available in \texttt{LALSuite} \cite{lalsuite} as the \texttt{SEOBNRv4HM\_PA} waveform approximant.}) which contains the modes, $(l, |m|)=(2,2),(2,1)$, $(3,3)$, $(4,4)$, and $(5,5)$~\cite{Cotesta:2018fcv,Mihaylov:2021bpf}. The model uses an effective-one-body approach combined with a post-adiabatic solution of the equations of motion to describe the inspiral-plunge waveform, $h_{l m}^\mathrm{insp-plunge}$. An accurate description of the merger is incorporated through calibration with NR simulations, as described in~\cite{Cotesta:2018fcv}, along with information for the merger and ringdown phases, from BH perturbation theory. The merger-ringdown waveform, $h_{l m}^\mathrm{merger-RD}$, is then stitched to inspiral-plunge waveform, $h_{l m}^\mathrm{insp-plunge}$ at a certain time $t = t^{\textrm{match}}_{l m}$, as
%
\begin{equation}
h_{l m}(t) = h_{l m}^\mathrm{insp-plunge}\, \theta(t_\mathrm{match}^{l m} - t) + h_{l m}^\mathrm{merger-RD}\,\theta(t-t_\mathrm{match}^{l m})\,,
\end{equation}
%
where $\theta(t)$ is the Heaviside step function. The merger-ringdown waveform is expressed as an exponentially damped sinusoid ~\citep{Bohe:2016gbl,Cotesta:2018fcv,Mihaylov:2021bpf}
%
\begin{equation}
\label{RD}
h_{l m}^{\textrm{merger-RD}}(t) = \nu \ \tilde{A}_{l m}(t)\ e^{i \tilde{\phi}_{l m}(t)} \ e^{-i \sigma_{l m 0}(t-t^{\textrm{match}}_{l m})},
\end{equation}
%
where
%
\begin{align}
\sigma_{l m 0} &= {\rm Re}(\sigma_{l m 0}) + i \, {\rm Im}(\sigma_{l m 0})\,,
\nonumber \\
&= \omega_{l m 0} - i / \tau_{l m 0}\,,
\end{align}
%
are the complex frequencies of the fundamental (n=0 overtone) QNMs of the remnant BH. The frequencies $\omega_{l m 0}$ and damping times $\tau_{l m 0}$ of the $(l, m)$ QNM can be read off from the real and imaginary parts of  $\sigma_{l m0}$ respetively. The functions $\tilde{A}_{l m}(t)$ and $\tilde{\phi}_{l m}(t)$ are defined in~\cite{Bohe:2016gbl,Cotesta:2018fcv}.

@@

\subsubsection{Bayesian formalism}
If we assume that a GW signal observed in detector data $d$ is accurately
described by our waveform model \pSEOB{}, we can infer the parameters of the
model, $\bm{\lambda}$, given the hypothesis $\mathcal{H}$, using the Bayes
theorem,
%
\begin{equation}
P(\bm{\lambda} \vert d, \mathcal{H}) =
\frac{P(\bm{\lambda} \vert \mathcal{H}) \, P(d \vert \bm{\lambda},\mathcal{H})}{P(d \vert \mathcal{H})}\,,
\label{eq:bayes}
\end{equation}
%
where $P(\bm{\lambda} \vert d, \mathcal{H})$ is the posterior probability distribution,
$P(\bm{\lambda} \vert \mathcal{H})$ the prior,
$P(d \vert \bm{\lambda},\mathcal{H})$ the likelihood, and
$P(d \vert \mathcal{H})$ the evidence.
%
The set of parameters, $\bm{\lambda}$ is the union between the GR waveform
model parameters~$\bm{\theta}$, detailed in Sec.~\ref{sec:review_pSEOB}, and
the non-GR parameter of $\ell$, i.e.,
%
\begin{equation}
\bm{\lambda} = \{\ell\} \cup \{\bm{\theta}\}.
\end{equation}
%
which, we recall, set the characteristic lengthscale in which deviations from
GR become relevant in each of the theories described in
Sec.~\ref{sec:review_theories}.

% As discussed in Secs.~\ref{sec:review_theories} and \ref{sec:theory_specific_qnm}, there is a single coupling
% constants in the cases of EdGB and dCS theories, but two in the cubic and
% quartic EFTofGR.
% %
% We also hold the other beyond-GR parameters, $\{\delta \omega_k^{(j)}\},$ and $\{\delta \tau_k^{(j)}\}$
% fixed to theory-specific predictions. Hence, our set of beyond--GR parameters only includes the parameter $\{\ell \}$.
% %

Assuming Gaussian and stationary noise, we can write the (log) likelihood function as,
%
\begin{equation}
\ln P(d \vert \bm{\lambda},\mathcal{H}) \propto
- \tfrac{1}{2}
\langle
d - h(\bm{\lambda}) \vert d - h(\bm{\lambda})
\rangle\,,
\end{equation}
%
with noise-weighted inner product $\langle \cdot | \cdot \rangle$ defined as,
%
\begin{equation}
\langle A | B \rangle =
\int_{f_{\rm low}}^{f_{\rm high}} \, \dd f \,
\frac{\tilde{A}^{\ast}(f) \tilde{B}(f) + \tilde{A}(f) \tilde{B}^{\ast}(f)}{S_{n}(f)}
\end{equation}
%
where $\tilde{A}(f)$ is the Fourier transform of $A(t)$, the asterisk denotes
complex conjugation and $S_{n}(f)$ is the power spectrum density of the
detector.
%
Assuming a specific prior distribution for our parameters, we stochastically
sample over the parameter space using a \texttt{LALInference} Markov-Chain
Monte Carlo algorithm~\cite{lalsuite}.
%
We subsequently marginalize over the remaining parameters to obtain the
posterior probability distribution function~(PDF) on $\ell$.
%
An example of this is shown in Fig. XX.

@@

If we find that our posterior obey both conditions~\eqref{eq:eft_bound}
and~\eqref{eq:parspec_bound}, we combine individual posteriors on $\ell$ from
each event.

Suppose we have a set of $N$ GW signals $\{D_{i}\}$, described by the ...
%
% When possible, we also combine the marginalized posterior distribution on
% $\ell$ for both event GW150914 and GW200129.
%
To do so, we employ one of the approaches of Ref.~\cite{Perkins:2021mhb} which
is relies on first obtaining a kernel density estimator (KDE) as an approximation for
the posterior distribution on $\ell$ for each individual events. Next, we multiply
these posteriors together to obtain the cumulative posterior.
%
(See also Refs.~\cite{DelPozzo:2011pg,Cardenas-Avendano:2019zxd,Carullo:2021dui}.)
%
This method can have difficulties in dealing with hard cut-offs in the
distribution, such as the one we impose at $\ell = 0$. We address this issue as
done in Ref.~\cite{Perkins:2021mhb}, by first doubling our number of samples as
$\{ s_i \} \cup \{ - s_i \}$, i.e., mirroring the samples across $\ell = 0$.
%
We then renormalize the associated KDE numerically in the range $\ell = [0,
\infty]$.
